---
title: "hastyDE versus limmaDE"
subtitle: "A focused simulation study for SpaceMosaic"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
    theme: flatly
    highlight: tango
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 10,
  fig.height = 5.8,
  out.width = "100%"
)

report_path <- normalizePath(knitr::current_input(dir = TRUE), mustWork = TRUE)
project_root <- normalizePath(file.path(dirname(report_path), ".."), mustWork = TRUE)

knitr::opts_knit$set(root.dir = project_root)
devtools::load_all(project_root, quiet = TRUE)
```

## Question

**Does limma's empirical-Bayes variance moderation improve the stability of
gene-level inference relative to the current matrix-based OLS method, while
leaving estimated effects and model residuals unchanged?**

`hastyDE()` is computationally efficient, but estimates each gene's residual
variance independently. This can be unstable in small spatial patches.
`limmaDE()` fits the same linear models and then uses `limma::eBayes()` to share
variance information across genes. The comparison therefore isolates the
effect of variance moderation without changing the underlying regression.

## Why simulation is the appropriate tool

Real data do not reveal which genes truly have an effect. Simulation gives us
known ground truth, allowing us to evaluate both null calibration and detection
power. The same simulated matrices and design are passed to both methods, so
any difference in standard errors or p-values can be attributed to empirical
Bayes moderation.

We examine two scenarios:

1. **Gaussian expression**, which directly matches the linear-model assumptions.
2. **Count-derived Pearson residuals**, which resembles the intended
   SpaceMosaic preprocessing. This is not a voom analysis: both methods receive
   the same residual matrix.

## Simulation design

We simulate one small patch containing 16 cells and 1,000 genes. The first 100
genes have a non-zero exposure effect; the other 900 are null. Gene-specific
variances are heterogeneous so that variance moderation has something meaningful
to estimate.

```{r simulate}
set.seed(18)

n_cells <- 16L
n_genes <- 1000L
n_de <- 100L
gene_names <- paste0("gene_", seq_len(n_genes))
is_de <- seq_len(n_genes) <= n_de
exposure <- as.numeric(scale(stats::rnorm(n_cells)))
df <- data.frame(exposure = exposure)

sigma <- exp(stats::rnorm(n_genes, mean = 0, sd = 0.7))
beta <- numeric(n_genes)
beta[is_de] <- 0.5 * sigma[is_de]
baseline <- stats::rnorm(n_genes, mean = 5, sd = 1)

y_gaussian <- outer(rep(1, n_cells), baseline) +
  outer(exposure, beta) +
  matrix(stats::rnorm(n_cells * n_genes), nrow = n_cells) *
  rep(sigma, each = n_cells)
colnames(y_gaussian) <- gene_names
rownames(y_gaussian) <- paste0("cell_", seq_len(n_cells))

library_size <- round(exp(stats::rnorm(n_cells, log(1000), 0.35)))
abundance <- exp(stats::rnorm(n_genes, log(0.02), 0.8))
count_beta <- numeric(n_genes)
count_beta[is_de] <- 0.12
mu <- outer(library_size, abundance) * exp(outer(exposure, count_beta))
y_counts <- matrix(
  stats::rnbinom(n_cells * n_genes, mu = as.vector(mu), size = 10),
  nrow = n_cells,
  dimnames = list(rownames(y_gaussian), gene_names)
)
y_pearson <- pearsonResiduals(y_counts, tot = library_size)
```

## Model comparison

For both scenarios, `limmaDE()` uses `trend = FALSE`: Pearson residuals are
already designed to remove the count mean-variance relationship, and the
Gaussian simulation has no imposed mean-variance trend. Robust empirical Bayes
estimation protects the hyperparameter estimates from unusually variable genes.

```{r fit-models}
fit_pair <- function(y) {
  list(
    hasty = hastyDE(y, df, return_residuals = TRUE),
    limma = limmaDE(
      y, df,
      return_residuals = TRUE,
      trend = FALSE,
      robust = TRUE
    )
  )
}

gaussian_fit <- fit_pair(y_gaussian)
pearson_fit <- fit_pair(y_pearson)
```

The maximum absolute discrepancies confirm that both approaches fit the same
coefficients and residuals:

```{r agreement, echo=FALSE}
agreement <- data.frame(
  Quantity = c("Gaussian effects", "Gaussian residuals",
               "Pearson effects", "Pearson residuals"),
  `Maximum absolute difference` = c(
    max(abs(gaussian_fit$hasty$effect - gaussian_fit$limma$effect)),
    max(abs(gaussian_fit$hasty$residuals - gaussian_fit$limma$residuals)),
    max(abs(pearson_fit$hasty$effect - pearson_fit$limma$effect)),
    max(abs(pearson_fit$hasty$residuals - pearson_fit$limma$residuals))
  ),
  check.names = FALSE
)
knitr::kable(agreement, digits = 3, row.names = FALSE)
```

## Results

```{r summarize, include=FALSE}
summarize_fit <- function(fits, scenario) {
  data.frame(
    Scenario = scenario,
    Method = c("hasty", "limma"),
    `False-positive rate` = c(
      mean(fits$hasty$p[!is_de, "exposure"] < 0.05),
      mean(fits$limma$p[!is_de, "exposure"] < 0.05)
    ),
    Power = c(
      mean(fits$hasty$p[is_de, "exposure"] < 0.05),
      mean(fits$limma$p[is_de, "exposure"] < 0.05)
    ),
    `Median SE` = c(
      median(fits$hasty$se[, "exposure"]),
      median(fits$limma$se[, "exposure"])
    ),
    check.names = FALSE
  )
}

summary_table <- rbind(
  summarize_fit(gaussian_fit, "Gaussian expression"),
  summarize_fit(pearson_fit, "Pearson residuals")
)
```

```{r summary-table, echo=FALSE}
knitr::kable(summary_table, digits = 3, row.names = FALSE)
```

The false-positive rate is the fraction of null genes with (p < 0.05).
Power is the corresponding fraction among truly non-null genes. These values
describe one realization and should not be interpreted as stable performance
estimates.

```{r comparison-plots, echo=FALSE, fig.height=7.2}
qq_coordinates <- function(p_values) {
  list(
    expected = -log10(stats::ppoints(length(p_values))),
    observed = -log10(sort(p_values))
  )
}

plot_comparison <- function(fits, scenario) {
  plot(
    fits$hasty$effect[, "exposure"],
    fits$limma$effect[, "exposure"],
    pch = 16, cex = 0.45,
    col = ifelse(is_de, "#D95F02", "#66666680"),
    xlab = "hastyDE effect", ylab = "limmaDE effect",
    main = paste(scenario, "effects")
  )
  abline(0, 1, lty = 2)

  plot(
    fits$hasty$se[, "exposure"], fits$limma$se[, "exposure"],
    pch = 16, cex = 0.45, col = "#377EB880",
    xlab = "hastyDE SE", ylab = "limmaDE moderated SE",
    main = paste(scenario, "standard errors")
  )
  abline(0, 1, lty = 2)

  hasty_qq <- qq_coordinates(fits$hasty$p[!is_de, "exposure"])
  limma_qq <- qq_coordinates(fits$limma$p[!is_de, "exposure"])
  qq_limit <- max(unlist(hasty_qq), unlist(limma_qq))
  plot(
    hasty_qq$expected, hasty_qq$observed,
    type = "l", col = "#1B9E77",
    xlim = c(0, qq_limit), ylim = c(0, qq_limit),
    xlab = "Expected -log10(p)", ylab = "Observed -log10(p)",
    main = paste(scenario, "null QQ plot")
  )
  lines(limma_qq$expected, limma_qq$observed, col = "#7570B3")
  abline(0, 1, lty = 2)
  legend("topleft", c("hasty", "limma"),
         col = c("#1B9E77", "#7570B3"), lty = 1, bty = "n")
}

par(mfrow = c(2, 3), mar = c(4, 4, 3, 1))
plot_comparison(gaussian_fit, "Gaussian")
plot_comparison(pearson_fit, "Pearson residual")
```

### How to interpret the panels

- **Effects:** points on the identity line confirm that empirical Bayes does
  not alter coefficients. Orange points are truly non-null genes; grey points
  are null genes.
- **Standard errors:** points above the identity line indicate that limma
  increased an unusually small SE; points below indicate shrinkage of an
  unusually large SE. This stabilization is the intended effect of moderation.
- **Null QQ plots:** the diagonal represents uniformly distributed null
  p-values. Curves above it suggest too many small p-values; curves below it
  indicate conservative inference.

## Small real-data example

### Question

**Across small data-driven spatial patches, how much does limma moderation
change uncertainty and gene prioritization relative to hastyDE?**

Unlike the simulation, real data provide no ground truth. This section therefore
does not ask which method is more accurate. It asks whether empirical-Bayes
moderation has a material practical effect when the model, cells, genes and
response transformation are held fixed.

We use the mouse-colon spatial transcriptomics dataset distributed with
SpaceMosaic. The analysis mirrors the immune-surveillance scenario in the main
vignette: patches are constructed among epithelial cells to retain local
variation in the number of immune cells in each cell's 100-nearest-neighbour
environment. Differential expression is then fitted separately within every
eligible patch.

### Patch construction and gene filtering

We first calculate immune density and a cellular-neighbourhood embedding on the
complete tissue. We then use `getPatches()` on all epithelial cells, using
spatial coordinates, immune density and neighbourhood context to create 100
candidate patches. Patch construction is completed before gene expression is
tested and uses no gene-level DE statistics, so it is independent of the
comparison between hastyDE and limmaDE.

Genes are filtered once across **all epithelial cells**, retaining genes
detected in at least 5% of them. This preserves a common gene universe selected
without using the predictor associations or their p-values. Library totals are
calculated from the original unfiltered count matrix, as required for Pearson
residuals.

```{r real-data-setup}
required_real_packages <- c("FNN", "ggplot2", "HDF5Array", "SpatialExperiment")
missing_real_packages <- required_real_packages[
  !vapply(required_real_packages, requireNamespace, logical(1), quietly = TRUE)
]
if (length(missing_real_packages) > 0L) {
  stop("Missing packages: ", paste(missing_real_packages, collapse = ", "))
}
suppressPackageStartupMessages(library(ggplot2))

spe <- HDF5Array::loadHDF5SummarizedExperiment(
  file.path("inst", "extdata", "mouse_colon")
)
counts <- SummarizedExperiment::assay(spe, "counts")
xy <- SpatialExperiment::spatialCoords(spe)
cell_type <- spe$tier1
is_epithelial <- cell_type == "Epithelial"
is_immune <- cell_type == "Immune"
Z <- embedCellNeighborhoods(
  mat = SingleCellExperiment::reducedDim(spe, "PCA"),
  xy = xy,
  ks = c(5, 50)
)

nn_all <- FNN::get.knn(as.matrix(xy), k = 100)$nn.index
immune_density <- rowSums(
  matrix(is_immune[nn_all], nrow = nrow(xy))
)

epithelial_index <- which(is_epithelial)
epithelial_xy <- xy[epithelial_index, , drop = FALSE]
detection_rate <- Matrix::rowMeans(counts[, is_epithelial, drop = FALSE] > 0)
keep_gene <- detection_rate >= 0.05

rownames(epithelial_xy) <- spe$cell_id[epithelial_index]
set.seed(123)
patch_fit <- getPatches(
  xy = epithelial_xy,
  X = immune_density[is_epithelial],
  npatches = 100,
  Z = Z[is_epithelial, , drop = FALSE],
  n_iters = 10,
  log_iters = TRUE,
  verbose = FALSE
)

patch_diagnostics <- getPatchDiagnostics(
  xy = epithelial_xy,
  patches = patch_fit,
  X = immune_density[is_epithelial]
)

eligible_patch <- patch_diagnostics$patch_diagnostics[
  patch_diagnostics$patch_diagnostics$n_cells >= 30 &
    is.finite(patch_diagnostics$patch_diagnostics$x_sd) &
    patch_diagnostics$patch_diagnostics$x_sd > 0, "patch"
]

selected_epithelial <- !is.na(patch_fit$patch) & patch_fit$patch %in% eligible_patch
selected_index <- epithelial_index[selected_epithelial]
selected <- seq_len(ncol(counts)) %in% selected_index

selected_totals <- Matrix::colSums(counts[, selected, drop = FALSE])
real_y <- t(counts[keep_gene, selected, drop = FALSE])
real_df <- data.frame(immune_density = immune_density[selected])
real_patch <- patch_fit$patch[selected_epithelial]

poly_df <- getPatchPolys(
  xy = epithelial_xy[selected_epithelial, , drop = FALSE],
  patch = real_patch,
  patch_data = patch_diagnostics$patch_diagnostics[
    patch_diagnostics$patch_diagnostics$patch %in% eligible_patch,]
)
```

Of 100 requested patches, `r length(unique(real_patch))` pass the eligibility
rule, containing `r sum(selected)` epithelial cells in total. Eligible patch
sizes range from `r min(table(real_patch))` to `r max(table(real_patch))` cells;
`r sum(keep_gene)` genes pass global filtering. Immune-neighbour counts range
from `r min(real_df$immune_density)` to `r max(real_df$immune_density)`.

```{r real-spatial-map, echo=FALSE, fig.height=5.2}
#| fig-width: 7
#| fig-height: 5.2
ggplot() +
    geom_point(data = xy[!is_epithelial, , drop = FALSE], aes(x = x, y = y),
        color = "grey", size = 0.1, alpha = 0.5) +
    geom_point(data = xy[is_epithelial, , drop = FALSE], aes(x = x, y = y),
        color = "blue", size = 0.1, alpha = 0.5) +
    geom_polygon(data = poly_df, aes(x = x, y = y, group = patch),
        fill = NA, color = "black", linewidth = 0.5, show.legend = FALSE) +
    coord_fixed() +
    theme_minimal() +
    labs(title = "Patches of epithelial cells",
         subtitle = "All outlined patches are included in the DE comparison") +
    theme(legend.position = "none")
```

### Same data and model, different variance inference

Both calls below compute the same patch-specific Pearson residuals. The only
difference is whether residual variances are estimated independently by OLS or
moderated across genes by limma.

```{r real-models}
hasty_timing <- system.time(real_hasty <- patchDE(
  y = real_y,
  df = real_df,
  patch = real_patch,
  method = "hasty",
  pearson = TRUE,
  tot = selected_totals,
  return_residuals = TRUE,
  verbose = FALSE
))
limma_timing <- system.time(real_limma <- patchDE(
  y = real_y,
  df = real_df,
  patch = real_patch,
  method = "limma",
  pearson = TRUE,
  tot = selected_totals,
  return_residuals = TRUE,
  verbose = FALSE
))

hasty_result <- real_hasty$de$immune_density
limma_result <- real_limma$de$immune_density
real_effect_difference <- max(
  abs(hasty_result$ests - limma_result$ests), na.rm = TRUE
)
real_residual_difference <- max(
  abs(real_hasty$residuals - real_limma$residuals), na.rm = TRUE
)
```

As expected for the same linear model, the maximum absolute effect difference
is `r format(real_effect_difference, scientific = TRUE, digits = 3)` and the
maximum residual difference is
`r format(real_residual_difference, scientific = TRUE, digits = 3)`.

```{r real-comparison-plots, echo=FALSE, fig.height=7.0}
hasty_effect <- as.vector(hasty_result$ests)
limma_effect <- as.vector(limma_result$ests)
hasty_se <- as.vector(hasty_result$ses)
limma_se <- as.vector(limma_result$ses)
hasty_p <- as.vector(hasty_result$pvals)
limma_p <- as.vector(limma_result$pvals)
finite_pair <- is.finite(hasty_effect) & is.finite(limma_effect) &
  is.finite(hasty_se) & is.finite(limma_se) &
  is.finite(hasty_p) & is.finite(limma_p)

par(mfrow = c(2, 2), mar = c(4, 4, 3, 1))
plot(
  hasty_effect[finite_pair], limma_effect[finite_pair],
  pch = 16, cex = 0.55, col = "#55555580",
  xlab = "hastyDE effect", ylab = "limmaDE effect",
  main = "Estimated effects"
)
abline(0, 1, lty = 2)

plot(
  hasty_se[finite_pair], limma_se[finite_pair],
  pch = 16, cex = 0.55, col = "#377EB880",
  xlab = "hastyDE SE", ylab = "limmaDE moderated SE",
  main = "Standard errors"
)
abline(0, 1, lty = 2)

plot(
  -log10(hasty_p[finite_pair]), -log10(limma_p[finite_pair]),
  pch = 16, cex = 0.55, col = "#7570B380",
  xlab = "hastyDE -log10(p)", ylab = "limmaDE -log10(p)",
  main = "Gene-level evidence"
)
abline(0, 1, lty = 2)

hasty_rank <- apply(
  hasty_result$pvals, 2, rank, ties.method = "average", na.last = "keep"
)
limma_rank <- apply(
  limma_result$pvals, 2, rank, ties.method = "average", na.last = "keep"
)
plot(
  as.vector(hasty_rank), as.vector(limma_rank),
  pch = 16, cex = 0.55, col = "#1B9E7780",
  xlab = "hastyDE p-value rank", ylab = "limmaDE p-value rank",
  main = "Gene prioritization"
)
abline(0, 1, lty = 2)
```

The effects remain fixed, whereas movement away from the identity line in the
other panels quantifies the practical impact of moderation. In particular,
ranking changes occur because limma stabilizes gene-wise uncertainty, not
because it changes the estimated association.

### Impact on nominal significance calls

Patch-level p-values are not adjusted for testing many genes across many
patches. Consequently, the threshold (p < 0.05) is used here only as a
descriptive reference point. These are **nominal calls**, not reportable DE
discoveries; multiplicity should be addressed at the final meta-analysis stage.

```{r real-significance-summary, echo=FALSE}
hasty_significant <- hasty_result$pvals < 0.05
limma_significant <- limma_result$pvals < 0.05
valid_tests <- is.finite(hasty_result$pvals) & is.finite(limma_result$pvals)

call_summary <- data.frame(
  Classification = c(
    "Significant with both methods",
    "hastyDE only",
    "limmaDE only",
    "Not significant with either method"
  ),
  `Gene-patch pairs` = c(
    sum(valid_tests & hasty_significant & limma_significant),
    sum(valid_tests & hasty_significant & !limma_significant),
    sum(valid_tests & !hasty_significant & limma_significant),
    sum(valid_tests & !hasty_significant & !limma_significant)
  ),
  check.names = FALSE
)
call_summary$Percent <- 100 * call_summary$`Gene-patch pairs` /
  sum(call_summary$`Gene-patch pairs`)
knitr::kable(call_summary, digits = 2, row.names = FALSE)
```

The following table focuses on the top 10 discordant pairs: cases where the selected DE
backend changes whether the nominal threshold is crossed. Pairs are ordered by
their strongest p-value, so the table emphasizes changes closest to the
evidence-bearing end of the results rather than the largest arbitrary rank
movements.

```{r real-discordant-table, echo=FALSE}
discordant <- valid_tests & xor(hasty_significant, limma_significant)
discordant_index <- which(discordant)
discordant_index <- discordant_index[
  order(
    pmin(
      hasty_result$pvals[discordant_index],
      limma_result$pvals[discordant_index]
    )
  )
]
shown_index <- head(discordant_index, 10)
pair_index <- arrayInd(shown_index, dim(hasty_result$pvals))

discordant_table <- data.frame(
  Gene = rownames(hasty_result$pvals)[pair_index[, 1]],
  Patch = colnames(hasty_result$pvals)[pair_index[, 2]],
  Effect = hasty_result$ests[shown_index],
  `hasty SE` = hasty_result$ses[shown_index],
  `limma SE` = limma_result$ses[shown_index],
  `hasty p` = hasty_result$pvals[shown_index],
  `limma p` = limma_result$pvals[shown_index],
  `Nominal call` = ifelse(
    hasty_significant[shown_index], "hastyDE only", "limmaDE only"
  ),
  check.names = FALSE
)
knitr::kable(discordant_table, digits = 4, row.names = FALSE)
```

Discordance arises only from variance moderation: the estimated effect is the
same, but the moderated standard error changes the corresponding test statistic
and p-value.

### Runtime

```{r real-runtime, echo=FALSE}
runtime_hasty <- hasty_timing[["elapsed"]]
runtime_limma <- limma_timing[["elapsed"]]

knitr::kable(
  data.frame(
    Method = c("hasty", "limma"),
    `Elapsed seconds` = c(runtime_hasty, runtime_limma),
    check.names = FALSE
  ),
  digits = 4,
  row.names = FALSE
)
```

These timings describe one execution across the eligible patches on the current
machine. They show practical overhead but are not a full benchmark.

## Conclusion and limitations

This toy study confirms the intended mechanism: `limmaDE()` preserves the
estimated effects and residuals while moderating uncertainty across genes. In
this realization, both methods show broadly similar null calibration, while
the changes in standard errors are clearly visible. The real-data example then
shows how those changes propagate to p-values and within-patch gene
prioritization across data-driven spatial patches, without implying that either
ranking is ground truth.

The study is deliberately small and explanatory. It does **not** establish that
one method is generally superior. A decision about the default SpaceMosaic
backend should rely on repeated simulations spanning patch size, sparsity,
effect size, variance heterogeneity and residual spatial autocorrelation.
