---
title: "SpaceMosaic Analysis Template"
author: "Your name"
date: "`r Sys.Date()`"
output:
  BiocStyle::html_document:
    toc: true
    toc_float: true
vignette: >
  %\VignetteIndexEntry{SpaceMosaic analysis template}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 8,
  fig.height = 6
)
library(SpaceMosaic)
library(SpatialExperiment)
library(ggplot2)
library(viridis)
```

# Overview

This template provides a dataset-independent starting point for a
**SpaceMosaic** analysis. Replace the marked inputs and thresholds with values
appropriate for the biological question and tissue geometry. The workflow is:

1. define the response-cell subset and design variable(s) ($X$);
2. encode the cellular microenvironment ($Z$);
3. identify and quality-control spatially contiguous patches;
4. estimate within-patch expression associations;
5. diagnose residual spatial autocorrelation for selected genes and patches;
6. borrow information among patches with similar contexts ($W$); and
7. interpret recurrent or context-specific effects in space.

Throughout the template, matrices are oriented as follows:

- expression: **cells x genes**;
- $X$: **cells x predictors** (a vector is also accepted by `getPatches()`);
- $Z$: **cells x neighborhood features**; and
- $W$: **patches x patch features**.

# 1. Data and analysis question

Load a `SpatialExperiment` object and identify the columns containing the
cell annotation and any other metadata needed for interpretation.

```{r load-data, eval=FALSE}
# Replace this with the loader for your dataset.
spe <- load_your_data()

stopifnot(inherits(spe, "SpatialExperiment"))
stopifnot("counts" %in% assayNames(spe))
stopifnot(nrow(spatialCoords(spe)) == ncol(spe))

# Stable cell identifiers are required as row names by getPatches().
xy <- as.matrix(spatialCoords(spe))
if (is.null(rownames(xy))) {
  rownames(xy) <- colnames(spe)
}

cell_type <- spe$cell_type       # CHANGE: broad cell annotation
cell_subtype <- spe$cell_subtype # CHANGE: optional detailed annotation
counts <- assay(spe, "counts")
```

Define one or more predictors that represent the biological exposure of
interest. Useful choices include distance to a cell population, local cell
abundance, neighborhood-mean expression, a spatial-domain indicator, or a
combination of predictors. Avoid deriving $X$ from the same genes later used as
responses unless that dependence is intentional.

```{r define-x-and-subset, eval=FALSE}
# Example: distance to the nearest cell of a selected explanatory population.
is_explanatory <- cell_type == "CHANGE_ME"
stopifnot(any(is_explanatory))

distance_to_explanatory <- FNN::get.knnx(
  data = xy[is_explanatory, , drop = FALSE],
  query = xy,
  k = 1
)$nn.dist[, 1]

# A log transform often reduces the leverage of long-tailed distances.
X <- data.frame(distance = log1p(distance_to_explanatory))

# Analyze only the response population relevant to the question.
use <- cell_type == "CHANGE_ME"
use <- use & stats::complete.cases(X)
stopifnot(sum(use) > 0)
```

# 2. Cellular-neighborhood embedding ($Z$)

`embedCellNeighborhoods()` averages a cell-level embedding over spatial
neighbor networks at multiple scales. Compute $Z$ on all cells before taking
the response-cell subset, so the representation retains the surrounding
tissue context. Supply `tissue` when the object contains multiple samples,
slides, or fields of view; this prevents neighborhoods from crossing tissue
boundaries.

```{r cellular-neighborhoods, eval=FALSE}
Z <- embedCellNeighborhoods(
  mat = reducedDim(spe, "PCA"),
  xy = xy,
  ks = c(5, 50),
  tissue = spe$sample_id # CHANGE; omit only for a single contiguous tissue
)

stopifnot(nrow(Z) == ncol(spe))
```

# 3. Patch identification and quality control

`getPatches()` favors spatially compact, contiguous regions that retain
within-patch variation in $X$ while remaining coherent in $Z$. For a
single-column $X$, `beta = 1` is a useful starting point. With $K$ predictors,
start near `beta = 1 / K`. Increase `alpha` to enforce stronger neighborhood
coherence, and inspect the result rather than treating the defaults as
universal.

```{r identify-patches, eval=FALSE}
set.seed(123)
patch_result <- getPatches(
  xy = xy[use, , drop = FALSE],
  X = X[use, , drop = FALSE],
  Z = Z[use, , drop = FALSE],
  npatches = 100,       # CHANGE for tissue area and response-cell abundance
  alpha = 0.5,          # weight of the Z-coherence penalty
  beta = 1 / ncol(X),   # weight of within-patch X diversity
  hunger_weight = 0.5, # balances patch size and X variation
  max_elongation = 4,
  mahal_radius = 3,
  n_candidates = 20,
  n_iters = 15,
  log_iters = TRUE
)
patch <- patch_result$patch

# Optional visual check of membership across iterations.
plotPatchIterations(
  xy = xy[use, , drop = FALSE],
  result = patch_result,
  iters = c(1, 5, 10, 15),
  ask = FALSE
)
```

`getPatchDiagnostics()` reports assignment coverage and summarizes patch size,
within-patch variation in `X`, spatial connectivity, and final membership
stability. These are general checks; RNA content, distance to an explanatory
population, tissue region, and sample identity can reveal additional
experiment-specific failure modes.

```{r patch-diagnostics, eval=FALSE}
diagnostics <- getPatchDiagnostics(
  xy = xy[use, , drop = FALSE],
  X = as.matrix(X[use, , drop = FALSE]),
  patches = patch_result
)
patch_diagnostics <- diagnostics$patch_diagnostics
diagnostics$assignment_summary

# CHANGE to the X column that should be inspected when X is multivariable.
x_sd_column <- if (ncol(X) == 1L) "x_sd" else paste0("x_sd_", colnames(X)[1])
patch_diagnostics$selected_x_sd <- patch_diagnostics[[x_sd_column]]
patch_diagnostics$connectivity <- factor(
  ifelse(
    patch_diagnostics$strict_component_fraction == 1,
    "Connected",
    "Fragmented"
  ),
  levels = c("Connected", "Fragmented")
)

# CHANGE after inspecting the diagnostic distributions and predictor scale.
min_cells <- 30
min_x_sd <- 0.1
min_stability <- 0.9

ggplot(patch_diagnostics, aes(n_cells, selected_x_sd)) +
  geom_point(aes(color = membership_stability, shape = connectivity), size = 2) +
  geom_vline(xintercept = min_cells, linetype = "dashed", color = "red") +
  geom_hline(yintercept = min_x_sd, linetype = "dashed", color = "red") +
  scale_color_viridis_c(limits = c(0, 1), na.value = "grey70") +
  scale_shape_manual(values = c(Connected = 16, Fragmented = 4)) +
  theme_minimal() +
  labs(
    title = "Core patch diagnostics",
    x = "Cells per patch",
    y = "Within-patch standard deviation of X",
    color = "Membership stability",
    shape = "Connectivity"
  )

keep <- with(
  patch_diagnostics,
  patch[
    n_cells >= min_cells &
      is.finite(selected_x_sd) & selected_x_sd >= min_x_sd &
      !is.na(membership_stability) &
      membership_stability >= min_stability &
      strict_component_fraction == 1
  ]
)
patch[!patch %in% keep] <- NA
```

The plot combines patch size, predictor variation, final stability, and
connectivity; crosses mark fragmentation at `strict_k`. Replace the
example thresholds after inspecting the dataset and the scale of `X`, then set
cells in excluded patches to `NA` as shown above.

Map the retained patches together with $X$ and the surrounding cells. Polygon
rows can carry any patch-level diagnostic or annotation supplied through
`patch_data`.

```{r patch-map, eval=FALSE}
poly_df <- getPatchPolys(
  xy = xy[use, , drop = FALSE],
  patch = patch,
  patch_data = patch_diagnostics
)

plot_df <- data.frame(
  x = xy[, 1],
  y = xy[, 2],
  use = use,
  predictor = X[, 1]
)

ggplot() +
  geom_point(
    data = plot_df[!plot_df$use, ],
    aes(x, y), color = "grey80", size = 0.2
  ) +
  geom_point(
    data = plot_df[plot_df$use, ],
    aes(x, y, color = predictor), size = 0.3
  ) +
  geom_polygon(
    data = poly_df,
    aes(x, y, group = patch),
    fill = NA, color = "black", linewidth = 0.3
  ) +
  scale_color_viridis_c() +
  coord_fixed() +
  theme_minimal()
```

# 4. Within-patch differential expression

`patchDE()` expects cells in rows and genes in columns. Therefore a conventional
genes-by-cells assay must be transposed. The rows of `df`, `y`, and `patch` must
refer to the same response cells in the same order.

For raw counts, `pearson = TRUE` computes Pearson residuals separately within
each patch and requires total counts per cell. Alternatively, transform or
normalize expression before calling `patchDE()` and include relevant technical
covariates in `df`. Do not use both strategies automatically; choose a model
appropriate for the assay and question.

```{r patch-de, eval=FALSE}
y <- t(counts[, use, drop = FALSE])
total_counts <- colSums(counts[, use, drop = FALSE])

de_design <- data.frame(
  X,
  # CHANGE: add known confounders, correctly subsetted to response cells.
  # batch = spe$batch,
  row.names = rownames(xy)
)
de_design <- de_design[use, , drop = FALSE]

stopifnot(
  nrow(y) == nrow(de_design),
  nrow(y) == length(patch),
  length(total_counts) == nrow(y)
)

de_result <- patchDE(
  y = y,
  df = de_design,
  patch = patch,
  pearson = TRUE,
  tot = total_counts
)
```

# 5. Residual spatial autocorrelation

Moran's I can be used as a diagnostic for spatial structure that remains after
the within-patch regression. Refit `patchDE()` only on the genes and patches of
interest, request the residuals, and pass one residual vector to `moranTest()`.
Filtering is explicit so that `y`, `df`, `patch`, and the spatial coordinates
remain aligned.

When `pearson = TRUE`, `total_counts` must remain the library totals calculated
from the original count matrix. Do not recompute them after selecting genes.
The residual matrix returned by `patchDE()` has the same rows and columns as
the already-filtered expression matrix supplied to it.

```{r residual-moran, eval=FALSE}
selected_patch <- names(which.max(table(patch))) # CHANGE to a patch of interest
selected_gene <- colnames(y)[1]                  # CHANGE to a gene of interest
selected_cells <- !is.na(patch) & patch == selected_patch

diagnostic_fit <- patchDE(
  y = y[selected_cells, selected_gene, drop = FALSE],
  df = de_design[selected_cells, , drop = FALSE],
  patch = patch[selected_cells],
  pearson = TRUE,
  tot = total_counts[selected_cells],
  return_residuals = TRUE,
  verbose = FALSE
)

selected_xy <- xy[use, , drop = FALSE][selected_cells, , drop = FALSE]

set.seed(123)
moran_result <- moranTest(
  residuals = diagnostic_fit$residuals,
  xy = selected_xy,
  patch = patch[selected_cells],
  k = 10,
  n_permutations = 999,
  alternative = "greater"
)
moran_result
```

Here, a small `p_value` with `alternative = "greater"` indicates that similar
residuals remain spatially clustered within the selected patch. This suggests
that the fitted model has not captured all spatial structure; it does not by
itself identify the missing biological or technical covariate.

# 6. Patch attributes ($W$) and local meta-analysis

`getPatchAttributes()` safely averages cell-level features within each retained
patch and sets patch identifiers as row names. $W$ can be based on $Z$, on
interpretable cell metadata, or on a combination of both. Its row names are
automatically aligned with the patch columns returned by `patchDE()`.

```{r meta-analysis, eval=FALSE}
W <- getPatchAttributes(
  Z = Z[use, , drop = FALSE],
  patch = patch
)

stopifnot(all(colnames(de_result[[1]]$ests) %in% rownames(W)))

meta_result <- patchMetaAnalysis(
  DEobj = de_result,
  W = W,
  k = 10,
  min_effect = 0,  # CHANGE if a minimum biologically relevant effect is known
  max_pval = 0.05,
  min_patches = 3
)
```

The result is keyed by modeled variable. For every variable it contains
posterior estimates, standard errors, p-values, and subgroup labels. Positive
and negative subgroup IDs mark connected regions in $W$-space with coherent
positive and negative effects; zero means that the patch did not belong to a
qualifying subgroup.

# 7. Result summaries and spatial validation

Rank genes using a criterion chosen in advance. The example below uses the
fraction of patches with an absolute posterior Z-score above 2. Effect sizes,
posterior p-values, subgroup membership, and biological relevance should also
be considered; very small p-values alone are not sufficient.

```{r result-summary, eval=FALSE}
variable <- names(meta_result)[1] # CHANGE when the design has several variables
meta_var <- meta_result[[variable]]
z_score <- meta_var$ests / meta_var$ses

gene_summary <- data.frame(
  gene = rownames(z_score),
  recurrent = rowMeans(abs(z_score) > 2, na.rm = TRUE),
  positive = rowMeans(z_score > 2, na.rm = TRUE),
  negative = rowMeans(z_score < -2, na.rm = TRUE)
)
gene_summary <- gene_summary[
  order(gene_summary$recurrent, decreasing = TRUE),
]
top_genes <- head(gene_summary$gene, 50)

heatmap(
  z_score[top_genes, , drop = FALSE],
  scale = "none"
)
```

Use `summarizeSubgroups()` to relate coherent effect regions to cell-level
metadata. Numeric variables are summarized as standardized enrichment scores;
categorical variables are represented by level-specific log2 enrichment.

```{r subgroup-summary, eval=FALSE}
cell_metadata <- data.frame(
  cell_type = cell_type[use],
  cell_subtype = cell_subtype[use],
  sample_id = spe$sample_id[use],
  predictor = X[use, 1]
)

subgroup_summary <- summarizeSubgroups(
  meta = meta_result,
  patch = patch,
  cellmeta = cell_metadata
)
head(subgroup_summary)
```

Finally, project representative effects back onto the tissue. Confirm that
their direction agrees with the predictor and is not driven by an unmodeled
sample, anatomical compartment, low-quality region, or segmentation artifact.

```{r spatial-validation, eval=FALSE}
gene <- top_genes[1]
poly_gene <- poly_df
poly_gene$z_score <- z_score[gene, as.character(poly_gene$patch)]

ggplot(poly_gene, aes(x, y, group = patch, fill = z_score)) +
  geom_polygon(color = "grey30", linewidth = 0.2) +
  scale_fill_gradient2(
    low = "blue", mid = "white", high = "red", midpoint = 0
  ) +
  coord_fixed() +
  theme_minimal() +
  labs(title = gene, fill = "Posterior Z")
```

For exploratory work, the interactive plotter can overlay cells and patches
and color patches by meta-analysis Z-scores or arbitrary patch annotations.

```{r interactive-plotter, eval=FALSE}
runInteractivePlotter(
  spe = spe[, use],
  patches = patch,
  metats = z_score,
  meaningful_vars = c("cell_type", "cell_subtype"),
  patch_data = patch_diagnostics
)
```

# 8. Reporting checklist

Record the definition and transformation of $X$, response-cell inclusion
criteria, construction of $Z$ and $W$, all `getPatches()` tuning parameters,
the random seed, patch-filtering thresholds, expression preprocessing,
covariates, meta-analysis parameters, and the number of retained cells and
patches. Report spatial validation and sensitivity analyses alongside the main
gene-level results.
