---
title: "Case Study: Human basal-cell carcinoma"
author: "Pere Moles"
date: "`r Sys.Date()`"
output:
  BiocStyle::html_document:
    toc: true
    toc_float: true
vignette: >
  %\VignetteIndexEntry{SpaceMosaic analysis template}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 8,
  fig.height = 6
)
library(SpaceMosaic)
library(SpatialExperiment)
library(ggplot2)
library(viridis)
```

# Overview

The workflow is:

1. define the response-cell subset and design variable(s) ($X$);
2. encode the cellular microenvironment ($Z$);
3. identify and quality-control spatially contiguous patches;
4. estimate within-patch expression associations;
5. borrow information among patches with similar contexts ($W$); and
6. interpret recurrent or context-specific effects in space.

Throughout the template, the input is a `SpatialExperiment` object with the following components:

- expression: assay **genes x cells**;
- $X$: column or columns in colData **cells x predictors**;
- $Z$: an embedding stored in reducedDims **cells x neighborhood features**; and
- $W$: patch-level attributes derived from $Z$ by `patchMetaAnalysis.spe()`
  (**patches x patch features**), returned alongside the meta-analysis results
  rather than stored back on `spe`.

The `.spe`-suffixed functions used below (`embedCellNeighborhoods.spe()`,
`getPatches.spe()`, `patchDE.spe()`, `patchMetaAnalysis.spe()`) are
`SpatialExperiment` wrappers: they read $X$, $Z$, expression, and patch
assignments directly from `spe`. `embedCellNeighborhoods.spe()` and
`getPatches.spe()` attach their results back onto `spe` and return it;
`patchDE.spe()` returns its differential-expression results alongside `spe`
(which also gains a `"residuals"` assay when `return_residuals = TRUE`); and
`patchMetaAnalysis.spe()` returns the meta-analysis results (`meta` and `W`,
plus `subgroups_summary` when requested) without `spe`, so intermediate
matrices rarely need to be built by hand.

# 1. Data and analysis question

Single-cell spatial transcriptomics (6k-plex CosMx SMI) dataset of human basal-cell carcinoma from https://zenodo.org/records/14330691 was preprocessed before the analysis.
The original Seurat object was converted to a `SpatialExperiment` object, and data was filtered for Patient A and FOVs 39 to 41. This consists of a nodular-ulcerated BCC sample.

```{r load-data}
spe <- readRDS(system.file("extdata", "cosmx_carcinoma.rds", package = "SpaceMosaic"))
```

We study the exposure of Cancer Cells near Fibroblasts.

```{r define-x-and-subset}
# Example: distance to the nearest cell of a selected explanatory population.
spe_explanatory <- spe[,spe$celltype == "Fibroblast"]

distance_to_explanatory <- FNN::get.knnx(
  data = spatialCoords(spe_explanatory),
  query = spatialCoords(spe),
  k = 1
)$nn.dist[, 1]

spe$distance <- log1p(distance_to_explanatory)
```

# 2. Cellular-neighborhood embedding ($Z$)

`embedCellNeighborhoods()` averages a cell-level embedding over spatial
neighbor networks at multiple scales. Compute $Z$ on all cells before taking
the response-cell subset, so the representation retains the surrounding
tissue context. Supply `tissue` when the object contains multiple samples,
slides, or fields of view; this prevents neighborhoods from crossing tissue
boundaries.

```{r cellular-neighborhoods}
spe <- embedCellNeighborhoods.spe(
  spe = spe,
  embedding = 'PCA',
  ks = c(5, 50),
  tissue = 'pat_fov'
)

head(reducedDim(spe, "Z")[, 1:5], 3)
```

# 3. Patch identification and quality control

`getPatches()` favors spatially compact, contiguous regions that retain
within-patch variation in $X$ while remaining coherent in $Z$.

```{r identify-patches}
spe_use <- spe[,spe$celltype == "Cancer.cells"]

set.seed(123)

X = 'distance'
spe_use <- getPatches.spe(
  spe_use,
  X = X,
  Z = 'Z',
  npatches = 100,       # CHANGE for tissue area and response-cell abundance
  alpha = 0.5,          # weight of the Z-coherence penalty
  beta = 1,   # weight of within-patch X diversity
  hunger_weight = 0.5, # balances patch size and X variation
  max_elongation = 4,
  mahal_radius = 3,
  n_candidates = 20,
  n_iters = 5,
  log_iters = TRUE
)

plotPatchIterations(
  xy = spatialCoords(spe_use),
  result = metadata(spe_use)$SpaceMosaic$patch_iterations,
  iters = 1:5,
  ask = FALSE
)
```


Patch diagnostics stored in metadata reports assignment coverage and summarizes patch size,
within-patch variation in `X`, spatial connectivity, and final membership
stability. These are general checks; RNA content, distance to an explanatory
population, tissue region, and sample identity can reveal additional
experiment-specific failure modes.

```{r patch-diagnostics}

patch_diagnostics <- metadata(spe_use)$SpaceMosaic$patch_diagnostics

# CHANGE to the X column that should be inspected when X is multivariable.
x_sd_column <- if (length(X) == 1L) "x_sd" else paste0("x_sd_", X[1])
patch_diagnostics$selected_x_sd <- patch_diagnostics[[x_sd_column]]
patch_diagnostics$connectivity <- factor(
  ifelse(
    patch_diagnostics$strict_component_fraction == 1,
    "Connected",
    "Fragmented"
  ),
  levels = c("Connected", "Fragmented")
)

# CHANGE after inspecting the diagnostic distributions and predictor scale.
min_cells <- 30
min_x_sd <- 0.1
min_stability <- 0.9

ggplot(patch_diagnostics, aes(n_cells, selected_x_sd)) +
  geom_point(aes(color = membership_stability, shape = connectivity), size = 2) +
  geom_vline(xintercept = min_cells, linetype = "dashed", color = "red") +
  geom_hline(yintercept = min_x_sd, linetype = "dashed", color = "red") +
  scale_color_viridis_c(limits = c(0, 1), na.value = "grey70") +
  scale_shape_manual(values = c(Connected = 16, Fragmented = 4)) +
  theme_minimal() +
  labs(
    title = "Core patch diagnostics",
    x = "Cells per patch",
    y = "Within-patch standard deviation of X",
    color = "Membership stability",
    shape = "Connectivity"
  )

keep <- with(
  patch_diagnostics,
  patch[
    n_cells >= min_cells &
      is.finite(selected_x_sd) & selected_x_sd >= min_x_sd &
      !is.na(membership_stability) &
      membership_stability >= min_stability &
      strict_component_fraction == 1
  ]
)

idx <- colData(spe_use)$patch %in% keep
colData(spe_use)$patch[!idx] <- NA

polys <- metadata(spe_use)$SpaceMosaic$patch_polys
idx_polys <- polys$patch %in% keep
metadata(spe_use)$SpaceMosaic$patch_polys <- polys[idx_polys, , drop = FALSE]
```

The plot combines patch size, predictor variation, final stability, and
connectivity; crosses mark fragmentation at `strict_k`. Replace the
example thresholds after inspecting the dataset and the scale of `X`, then set
cells in excluded patches to `NA` as shown above.

Overlay the retained patch boundaries on the per-cell map of $X$ to check
visually that patches trace coherent regions of the exposure variable rather
than cutting across it.

```{r patch-map}
library(ggspavis)

spe$distance_use <- ifelse(
  spe$celltype == "Cancer.cells",
  spe$distance,
  NA
)

plotCoords(spe,annotate = 'distance_use',in_tissue=NULL, y_reverse = FALSE)  +
  geom_polygon(
    data = metadata(spe_use)$SpaceMosaic$patch_polys,
    aes(x, y, group = patch),
    fill = NA, color = "black", linewidth = 0.3
  ) +
  scale_color_viridis_c() +
  coord_fixed() +
  theme_minimal()
```

# 4. Within-patch differential expression

`patchDE.spe()` wraps `patchDE()` for a `SpatialExperiment`: it takes the
assay named by `assay` (genes x cells), transposes it internally, builds the
predictor data frame from `predictor_cols` in `colData(spe)`, and reads patch
assignments from `colData(spe)[[patch_column]]`; cells with a missing patch
assignment are dropped automatically.

For raw counts, `pearson = TRUE` computes Pearson residuals separately within
each patch and requires `tot` to name a column in `colData(spe)` holding
per-cell total counts. Alternatively, transform or normalize expression
before calling `patchDE.spe()` and include relevant technical covariates in
`predictor_cols`.

```{r patch-de}
de_result <- patchDE.spe(
  spe = spe_use,
  predictor_cols = 'distance',
  method = "limma",
  pearson = FALSE,
  tot = 'total',
  return_residuals = TRUE,
  resid_mse = FALSE
)
```

# 5. Borrowing information across patches ($W$)

`patchMetaAnalysis.spe()` derives patch-level attributes ($W$) by averaging
the embedding named by `embedding_name` (default `"Z"`, taken from
`reducedDims(spe)`) within each patch, then borrows information across
patches with similar $W$ by running `patchMetaAnalysis()` on the per-patch
fits produced by `patchDE.spe()`. Set `summarize_subgroups = TRUE` to also
summarize effects within subgroups of patches, optionally annotated with
`cellmeta_cols` taken from `colData(spe)`.

```{r patch-meta}
meta_result <- patchMetaAnalysis.spe(
  patchDE_result = de_result,
  summarize_subgroups = TRUE,
  cellmeta_cols = c("celltype", "distance")
)
```



# 6. Result summaries and spatial validation

Rank genes using a criterion chosen in advance. The example below uses the
fraction of patches with an absolute posterior Z-score above 2. Effect sizes,
posterior p-values, subgroup membership, and biological relevance should also
be considered; very small p-values alone are not sufficient.

```{r result-summary}
z_score <- meta_result$meta$distance$ests/ meta_result$meta$distance$ses

gene_summary <- data.frame(
  gene = rownames(z_score),
  recurrent = rowMeans(abs(z_score) > 2, na.rm = TRUE),
  positive = rowMeans(z_score > 2, na.rm = TRUE),
  negative = rowMeans(z_score < -2, na.rm = TRUE)
)
gene_summary <- gene_summary[
  order(gene_summary$recurrent, decreasing = TRUE),
]
top_genes <- head(gene_summary$gene, 50)

heatmap(
  z_score[top_genes, , drop = FALSE],
  scale = "none"
)
```


`return_residuals = TRUE` stored the within-patch residuals as a
`"residuals"` assay on `de_result$spe`, which `moranTest.spe()` reads
directly, so gene selection is done by subsetting `de_result$spe` rather than
indexing a separate matrix.

```{r}

moran_test <- moranTest.spe(
  de_result$spe[top_genes, ],
  k = 20,
  n_permutations = 20
)

head(moran_test[order(moran_test$observed, decreasing = TRUE),])
```

Finally, project representative effects back onto the tissue. Confirm that
their direction agrees with the predictor and is not driven by an unmodeled
sample, anatomical compartment, low-quality region, or segmentation artifact.

```{r spatial-validation}
poly_df <- metadata(spe_use)$SpaceMosaic$patch_polys
gene <- top_genes[1]
poly_gene <- poly_df
poly_gene$z_score <- z_score[gene, ][as.character(poly_gene$patch)]

ggplot(poly_gene, aes(x, y, group = patch, fill = z_score)) +
  geom_polygon(color = "grey30", linewidth = 0.2) +
  scale_fill_gradient2(
    low = "blue", mid = "white", high = "red", midpoint = 0
  ) +
  coord_fixed() +
  theme_minimal() +
  labs(title = gene, fill = "Posterior Z")

```