---
title: "Patch diagnostics"
author: "Peiying Cai"
date: "`r Sys.Date()`"
output:
  BiocStyle::html_document:
    toc: true
    toc_float: true
    code_folding: hide
vignette: >
  %\VignetteIndexEntry{Patch diagnostics}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 8,
  fig.height = 6
)
library(SpaceMosaic)
library(HDF5Array)
library(SpatialExperiment)
library(ggplot2)
library(viridis)
library(patchwork)
```

# Introduction

`getPatchDiagnostics()` reports, for every patch produced by `getPatches()`, its size, its within-patch variation in the design variable `X`, whether its cells form one connected physical region, and (when iteration logs are available) the extent to which cell-to-patch assignments were still changing near the end of the algorithm.

This vignette uses the mouse colon dataset to show:

- how to run `getPatches()` and `getPatchDiagnostics()` together,
- what each diagnostic column means and how to read it,
- how to trace a diagnostic outlier (for example, a fragmented patch) back to its spatial location,
- how the diagnostics change across alternative `getPatches()` settings, to help choose `npatches`, and
- how to turn diagnostics into an investigative filtering step.

Differential expression and meta-analysis on the retained patches are covered separately in `vignette("mouse_colon")`.

# Setup: patches and diagnostics

We reuse the "Immune Surveillance" scenario from `vignette("mouse_colon")`: *how does the transcriptional state of the colon epithelium change as a function of its proximity to immune cells?* Epithelial cells are the analysis unit, and `X` is the number of immune cells among each cell's 100 nearest neighbors.

```{r data-load, fig.cap="Spatial distribution of epithelial and immune cells in the mouse colon sample. All other cell types are grey.", fig.align = 'center'}
h5_path <- system.file("extdata", "mouse_colon", package = "SpaceMosaic")
spe <- HDF5Array::loadHDF5SummarizedExperiment(h5_path)
rownames(spatialCoords(spe)) <- spe$cell_id
xy <- spatialCoords(spe)
cell_types <- spe$tier1

# Spatial distribution of epithelial and immune cells
coords <- data.frame(xy, cell_types = as.character(cell_types))
ggplot(coords, aes(x, y, color = ifelse(cell_types %in% c("Epithelial", "Immune"),
                   cell_types, "Other"))) +
  geom_point(size = 0.1) +
  coord_fixed() +
  theme_classic() +
  scale_color_manual(values = c(
    Epithelial = "blue",
    Immune = "orange",
    Other = "grey80")) +
  labs(color = "Cell Type") +
  guides(color = guide_legend(override.aes = list(size = 3)))
```

```{r embedding-and-predictor, class.source = "fold-show"}
# Compute Cellular Neighborhoods (Z)
Z <- embedCellNeighborhoods(
  mat = reducedDim(spe, "PCA"),
  xy = xy,
  ks = c(5, 50)
)

# Global setup
is_epithelial <- cell_types == "Epithelial"
is_immune <- cell_types == "Immune"

k_neighbors <- 100
nn_indices <- FNN::get.knn(data = as.matrix(xy), k = k_neighbors)$nn.index
X <- apply(nn_indices, 1, function(row) sum(is_immune[row]))
use <- is_epithelial
```

```{r patches, class.source = "fold-show"}
# Construct patches
set.seed(123)
patches_res <- getPatches(
  xy = spatialCoords(spe)[use, ],
  X = X[use],
  Z = Z[use, ],
  npatches = 100,
  n_iters = 15,
  log_iters = TRUE
)
patches <- patches_res$patch

# Patch diagnostics
diagnostics <- getPatchDiagnostics(
  xy = as.matrix(spatialCoords(spe)[use, , drop = FALSE]),
  X = X[use],
  patches = patches_res
)
names(diagnostics)
```

`getPatchDiagnostics()` returns a list with three data frames:

- `patch_diagnostics`: one row per patch, describing its size, within-patch variation in `X`, spatial connectivity, and membership stability (only when `getPatches(log_iters = TRUE)`).
- `assignment_summary`: one row describing coverage over all analyzed cells, such as how many were assigned to a patch and the neighbor counts used to evaluate connectivity.
- `connectivity_curve`: one row per patch per tested `k` neighbors, used below to show spatial fragmentation.

<details id="connectivity-details">
<summary>Details: how spatial connectivity is measured (click to expand)</summary>

`getPatches()` checks contiguity on a symmetric 10-nearest-neighbor (10-NN)
graph after every iteration and removes cells outside the largest connected
component of each patch. The diagnostics apply the same graph definition over
a range of neighborhood sizes: `strict_k = 5` by default for a deliberately
local check, and values up to `k = 10` by default. For each value of `k`, the
graph is first built among all cells supplied to `getPatchDiagnostics()`. An
edge is then retained only when both cells have the same non-missing patch
label. A chain of retained edges connects two cells, and a connected component
is the complete set of patch cells reachable through such chains.

The `largest-component fraction` is the size of a patch's biggest connected component divided by its total size (`n_cells`). It is 1 when the whole patch is one piece, and lower when part of it sits in smaller, disconnected pieces. It depends on `k`: a small `k` is a strict test, a larger `k` makes full connectivity easier to reach. `getPatchDiagnostics()` evaluates it across a range of `k` and reports it three ways: `connectivity_curve` (the fraction at every tested `k`), `strict_component_fraction` (the fraction at one small, fixed `strict_k`), and `min_connectivity_k` (the smallest `k` at which the fraction first reaches 1; `NA` if it never does, `0` for single-cell patches).

</details>

```{r patch-diagnostics-table, class.source = "fold-show"}
patch_data <- diagnostics$patch_diagnostics
head(patch_data)
```

`n_cells` is the number of cells assigned to the patch, and `x_sd` is the standard deviation of `X` within it. `strict_component_fraction` and `min_connectivity_k` summarize spatial connectivity (see [details above](#connectivity-details)). `membership_stability` is the fraction of the patch's final cells that were assigned to it in the preceding iteration. The rest of this vignette interprets each of these in turn.

```{r assignment-summary-table, class.source = "fold-show"}
diagnostics$assignment_summary
```

Here `r diagnostics$assignment_summary$n_unassigned_cells` of `r diagnostics$assignment_summary$n_analyzed_cells` epithelial cells (`r sprintf("%.1f%%", 100 * diagnostics$assignment_summary$fraction_unassigned)`) were not assigned to any of the `r diagnostics$assignment_summary$n_nonempty_patches` non-empty patches. Connectivity was evaluated up to `r diagnostics$assignment_summary$connectivity_k` neighbors (`connectivity_k`), using `r diagnostics$assignment_summary$strict_connectivity_k` neighbors (`strict_connectivity_k`) for the strict fraction below.

```{r connectivity-curve-table, class.source = "fold-show"}
head(diagnostics$connectivity_curve)
```

For each patch, this reports the largest connected component's size divided by the patch's total size, across every tested `k` from 1 up to `connectivity_k`. Here `k` is the number of spatial neighbors used to build the connectivity graph. Traced across `k`, these fractions form the connectivity curve.

# Overview of diagnostics results

We start with a scatter plot of cells per patch and the standard deviation of `X` within the patch, colored by membership stability and shaped by connectivity.

```{r core-diagnostics, fig.width=8, fig.height=6, fig.cap="Cells per patch versus within-patch SD of the immune-neighbor predictor. Colored by membership stability; shape marks connectivity."}
patch_data$connectivity <- factor(
  ifelse(patch_data$strict_component_fraction == 1, "Connected", "Fragmented"),
  levels = c("Connected", "Fragmented")
)

ggplot(patch_data, aes(x = n_cells, y = x_sd)) +
  geom_point(aes(color = membership_stability, shape = connectivity), size = 2) +
  scale_color_viridis_c(limits = c(0, 1), na.value = "grey70") +
  scale_shape_manual(values = c(Connected = 16, Fragmented = 4)) +
  theme_minimal() +
  labs(
    x = "Cells per patch",
    y = "Within-patch SD of immune-neighbor count",
    color = "Membership stability",
    shape = "Connectivity"
  )
```

Points near the bottom-left are small and nearly constant in `X`. Relative to
larger patches with more variation, they generally provide less information
for estimating a within-patch association with `X`. Dark, low-stability points
identify patches whose membership was still changing in the final update.
Crosses mark patches split into multiple spatial components at `strict_k`.

There is no universal cutoff for any of these values. In general, a good patch has enough cells and enough within-patch variation in `X`, is not so large that it loses spatial resolution, and forms one connected piece.

# Patch size and within-patch variation

```{r patch-polys}
# getPatchPolys() returns each patch's convex-hull vertices and joins the
# supplied patch-level diagnostics to those vertices for plotting.

poly_df <- getPatchPolys(
  xy = spatialCoords(spe)[use, ],
  patch = patches,
  patch_data = patch_data
)
```

```{r size-variation-map, fig.width=18, fig.height=6, fig.cap="Patch maps: predictor, within-patch variation, and size. Left: cells colored by X, with patch outlines. Middle: patches filled by within-patch SD of X. Right: patch size, highlighting the three smallest and three largest patches by n_cells."}
sel <- as.data.frame(xy)[use, ]
sel$patch <- patches
sel$X <- X[use]

# The three smallest and three largest patches by n_cells, highlighted below.
patch_centroids <- aggregate(cbind(x, y) ~ patch, data = sel, FUN = mean)
n_label <- 3
size_order <- patch_data$patch[order(patch_data$n_cells)]
extreme_patches <- c(head(size_order, n_label), tail(size_order, n_label))
label_df <- merge(
  patch_centroids[patch_centroids$patch %in% extreme_patches, ],
  patch_data[patch_data$patch %in% extreme_patches, c("patch", "n_cells")],
  by = "patch"
)

# Left: the predictor, with patch outlines
p_predictor <- ggplot() +
  geom_point(data = sel, aes(x, y, color = X), size = 0.4) +
  geom_polygon(data = poly_df, aes(x, y, group = patch), fill = NA, color = "#39FF14", linewidth = 0.7) +
  coord_fixed() + scale_color_viridis_c(option = "plasma") + theme_void() +
  labs(color = "X (immune neighbors)", x = NULL, y = NULL)

# Middle: within-patch variation
p_variation <- ggplot() +
  geom_polygon(data = poly_df, aes(x, y, group = patch, fill = x_sd), color = "black", linewidth = 0.7) +
  coord_fixed() + scale_fill_gradient(low = "white", high = "yellow") + theme_void() +
  labs(fill = "x_sd", x = NULL, y = NULL)

# Right: patch size
p_size <- ggplot() +
  geom_polygon(data = poly_df, aes(x, y, group = patch), fill = NA, color = "grey80", linewidth = 0.2) +
  geom_polygon(
    data = subset(poly_df, patch %in% extreme_patches),
    aes(x, y, group = patch, fill = n_cells), color = "black", linewidth = 0.6
  ) +
  geom_text(data = label_df, aes(x, y, label = n_cells), fontface = "bold", size = 3) +
  coord_fixed() + scale_fill_gradient(low = "white", high = "yellow") + theme_void() +
  labs(fill = "n_cells", x = NULL, y = NULL)

p_predictor + p_variation + p_size
```

The middle panel shows where the immune-neighbor predictor varies within a
patch, whereas the right panel locates the most extreme patch sizes. These
maps should be read descriptively: a low `x_sd` patch may be unsuitable for a
regression on `X`, and an unusually small or large patch may merit inspection,
but neither feature alone establishes that a patch is invalid.

# Connectivity and stability

## Spatial connectivity

This section examines the connectivity output from `getPatchDiagnostics()`
(see the [details above](#connectivity-details)).

```{r connectivity-curve-setup, class.source = "fold-show"}
# Extract strict_k, the fixed neighbor count used for strict_component_fraction calculation
strict_k <- diagnostics$assignment_summary$strict_connectivity_k

# Attach that connected/fragmented label
connectivity_curve <- diagnostics$connectivity_curve
connectivity_curve$connectivity <- patch_data$connectivity[
  match(connectivity_curve$patch, patch_data$patch)]
table(patch_data$connectivity)

# Pick the patch with the lowest strict_component_fraction
frag_patch <- patch_data$patch[which.min(patch_data$strict_component_fraction)]
```

With `strict_k` = `r strict_k`, `r sum(patch_data$connectivity == "Fragmented")` of the `r nrow(patch_data)` patches are flagged as fragmented.

```{r connectivity-curve-plot, fig.width=7, fig.height=4.5, fig.cap="Connectivity curves for every patch. Colored by whether the patch is fragmented at strict_k; the dashed line marks strict_k, and the black line is the worked example discussed below."}
ggplot(connectivity_curve, aes(k, component_fraction, group = patch)) +
  geom_line(aes(color = connectivity), alpha = 0.5, linewidth = 0.3) +
  geom_line(data = subset(connectivity_curve, patch == frag_patch), color = "black", linewidth = 1.2) +
  geom_vline(xintercept = strict_k, linetype = "dashed", color = "grey30") +
  scale_color_manual(values = c(Connected = "grey80", Fragmented = "#D55E00")) +
  scale_x_continuous(breaks = seq_len(max(connectivity_curve$k))) +
  theme_classic() +
  labs(x = "Number of spatial neighbors (k)", y = "Largest-component fraction", color = "Connectivity")
```

The black curve, which has the lowest `strict_component_fraction`, reaches 1
at `k = ` `r patch_data$min_connectivity_k[patch_data$patch == frag_patch]`. All patches reach 1
by the default maximum of `k = 10`, as expected after the final 10-NN
contiguity check in `getPatches()`.

```{r fragmented-patch-example, class.source = "fold-show"}
patch_data[patch_data$patch == frag_patch, c("patch", "n_cells", "strict_component_fraction", "min_connectivity_k")]
```

Patch `r frag_patch` has the lowest `strict_component_fraction`, and close to
half of its `r patch_data$n_cells[patch_data$patch == frag_patch]` cells fall
outside the largest component at `strict_k` = `r strict_k`. In the zoomed map,
the cells occupy one local region but form two groups under the
`r strict_k`-NN graph. Its `min_connectivity_k` of
`r patch_data$min_connectivity_k[patch_data$patch == frag_patch]` means that the
groups are nevertheless close enough to become connected when each cell can
link to one additional neighbor. In this sense, the split is local and resolves
after only a small increase in `k`.

```{r fragmented-patch-zoom, fig.width=11, fig.height=5, fig.cap="Zooming in on the split patch. Left: epithelial cells colored by X, with the highlighted patch outlined in green. Right: the same patch zoomed in, with its cells in orange and all other cells in grey."}
cell_idx <- !is.na(sel$patch) & sel$patch == frag_patch
poly_frag <- subset(poly_df, patch == frag_patch)
bbox <- apply(sel[cell_idx, c("x", "y")], 2, range)
pad <- 0.15 * (bbox[2, ] - bbox[1, ])

p_context <- ggplot() +
  geom_point(data = sel, aes(x, y, color = X), size = 0.1) +
  geom_polygon(data = poly_frag, aes(x, y, group = patch), fill = NA, color = "#39FF14", linewidth = 0.8) +
  coord_fixed() + scale_color_viridis_c(option = "plasma") + theme_minimal() + labs(x = NULL, y = NULL)

p_zoom <- ggplot() +
  geom_point(data = as.data.frame(xy), aes(x, y), color = "grey85", size = 1) +
  geom_point(data = sel[cell_idx, ], aes(x, y), color = "#D55E00", size = 1) +
  geom_polygon(data = poly_frag, aes(x, y, group = patch), fill = NA, color = "black", linewidth = 0.6) +
  coord_fixed(xlim = bbox[, "x"] + c(-1, 1) * pad["x"], ylim = bbox[, "y"] + c(-1, 1) * pad["y"]) +
  theme_minimal() + labs(x = NULL, y = NULL)

p_context + p_zoom
```

In the zoomed panel, orange points are epithelial cells assigned to the selected
patch. Grey points are all other cells in the sample. They may be epithelial
cells assigned to another patch or left unassigned, as well as non-epithelial
cells.

If a patch has a low `strict_component_fraction` and a comparatively large
`min_connectivity_k`, its pieces remain separate under a local graph and warrant
closer spatial inspection. For the default output of `getPatches()`, however,
the final orphan-removal step already retains only the largest component at
`k = 10`. The default diagnostics should therefore reach 1 by `k = 10`.

```{r spatial-connectivity-map, fig.width=8, fig.height=6, fig.cap="Spatial map of patch connectivity. Epithelial cells are grey; patch outlines are colored by strict_component_fraction."}
ggplot() +
  geom_point(data = as.data.frame(xy)[use, ], aes(x, y), color = "grey85", size = 0.1) +
  geom_polygon(data = poly_df, aes(x, y, group = patch, color = strict_component_fraction), fill = NA, linewidth = 0.5) +
  coord_fixed() + scale_color_viridis_c(limits = c(0, 1)) + theme_minimal() +
  labs(color = "strict_component_fraction", x = NULL, y = NULL)
```


## Membership stability

`membership_stability` is the fraction of a patch's final cells that were assigned to the same patch in the preceding iteration; it requires `getPatches(log_iters = TRUE)` and is `NA` otherwise. Values near 1 indicate little membership change in the final update. Lower values indicate that membership was still being reshuffled.

# Comparing different numbers of patches

`npatches` controls an important trade-off. Fewer requested patches generally
produce larger spatial units, whereas more requested patches generally produce
smaller units with finer spatial resolution. The effects on within-patch
variation, connectivity, and assignment coverage are data-dependent. We compare
three settings on the same data, predictor, and embedding, using fewer
iterations to keep this comparison quick.

```{r grid-setup, class.source = "fold-show"}
xy_use <- spatialCoords(spe)[use, , drop = FALSE]
X_use <- X[use]
Z_use <- Z[use, , drop = FALSE]

npatches_grid <- c(50L, 100L, 150L)

# Run getPatches() + getPatchDiagnostics() once per npatches setting
fits <- vector("list", length(npatches_grid))
diags <- vector("list", length(npatches_grid))
names(fits) <- names(diags) <- as.character(npatches_grid)

for (i in seq_along(npatches_grid)) {
  set.seed(123)
  fits[[i]] <- getPatches(
    xy = xy_use, X = X_use, Z = Z_use,
    npatches = npatches_grid[i], n_iters = 5, log_iters = TRUE
  )
  diags[[i]] <- getPatchDiagnostics(xy = xy_use, X = X_use, patches = fits[[i]])
}

# Tag npatches
add_npatches <- function(df, npatches) {df$npatches <- npatches; df}

# Stack results
stack_tables <- function(one_result_per_setting) {
  do.call(rbind, Map(add_npatches, one_result_per_setting, npatches_grid))
}

patch_data_grid <- stack_tables(lapply(diags, `[[`, "patch_diagnostics"))
connectivity_curve_grid <- stack_tables(lapply(diags, `[[`, "connectivity_curve"))
assignment_summary_grid <- stack_tables(lapply(diags, `[[`, "assignment_summary"))
poly_df_grid <- stack_tables(Map(
  function(fit, diag) getPatchPolys(xy = xy_use, patch = fit$patch, patch_data = diag$patch_diagnostics),
  fits, diags
))
```

The spatial view shows how patch size decreases as `npatches` increases:

```{r grid-spatial, fig.width=13, fig.height=5, fig.cap="Patch outlines across npatches settings. Epithelial cells are colored by the immune-neighbor predictor; green lines are patch outlines."}
ggplot() +
  geom_point(data = data.frame(xy_use, immune_neighbors = X_use), aes(x, y, color = immune_neighbors), size = 0.05) +
  geom_polygon(data = poly_df_grid, aes(x, y, group = patch), fill = NA, color = "#39FF14", linewidth = 0.5) +
  facet_wrap(~npatches, nrow = 1) + coord_fixed() +
  scale_color_viridis(option = "plasma") + theme_minimal() +
  labs(color = "Immune cells\namong 100-NN", x = NULL, y = NULL)
```

As `npatches` increases, patches visibly shrink and multiply. Whether the gain
in spatial resolution is worth the accompanying change in information per
patch depends on how the diagnostics below respond.

Density plots compare the size/variation trade-off across settings:

```{r grid-distributions, fig.width=10, fig.height=4, fig.cap="Distributions across different numbers of patches. Color and line type denote the requested number of patches; panels show cells per patch, membership stability, and the SD of the immune-neighbor predictor."}
metrics <- c(n_cells = "Cells per patch", x_sd = "SD of immune-neighbor count", membership_stability = "Membership stability")
distribution_data <- do.call(rbind, Map(function(col, label) {
  data.frame(npatches = patch_data_grid$npatches, metric = label, value = patch_data_grid[[col]])
}, names(metrics), metrics))
distribution_data$npatches <- factor(distribution_data$npatches, levels = npatches_grid)

ggplot(distribution_data, aes(value, color = npatches, linetype = npatches)) +
  geom_density(linewidth = 1) +
  facet_wrap(~metric, scales = "free", ncol = 3) +
  theme_minimal() +
  labs(x = NULL, y = "Density", color = "npatches")
```

The size distributions shift as expected: requesting fewer patches produces
larger patches. The `x_sd` and stability panels show whether that change also
alters within-patch predictor contrast or final-iteration membership for this
dataset. Compare the full distributions rather than relying on a single patch
or a universal cutoff.

The same trade-offs can also be viewed patch by patch:

```{r grid-diagnostics, fig.width=11, fig.height=5, fig.cap="Patch diagnostics across different numbers of patches. Cells per patch versus within-patch SD of the immune-neighbor predictor; color shows membership stability and shape shows connectivity."}
patch_data_grid$connectivity <- factor(
  ifelse(patch_data_grid$strict_component_fraction == 1, "Connected", "Fragmented"),
  levels = c("Connected", "Fragmented")
)
ggplot(patch_data_grid, aes(n_cells, x_sd)) +
  geom_point(aes(color = membership_stability, shape = connectivity), size = 2) +
  facet_wrap(~npatches, scales = "free_x") +
  scale_color_viridis_c(limits = c(0, 1), na.value = "grey70") +
  scale_shape_manual(values = c(Connected = 16, Fragmented = 4)) +
  theme_minimal() +
  labs(x = "Cells per patch", y = "Within-patch SD of immune-neighbor count",
       color = "Membership stability", shape = "Connectivity")
```


Connectivity curves, faceted the same way, compare local connectivity across settings:

```{r grid-connectivity, fig.width=11, fig.height=4, fig.cap="Connectivity curves across npatches settings. Grey lines are individual patches; the blue line is the median."}
ggplot(connectivity_curve_grid, aes(k, component_fraction, group = patch)) +
  geom_line(color = "grey60", alpha = 0.25, linewidth = 0.3) +
  stat_summary(aes(group = 1), fun = median, geom = "line", color = "#0072B2", linewidth = 1.1) +
  geom_vline(xintercept = strict_k, linetype = "dashed", color = "grey30") +
  facet_wrap(~npatches, nrow = 1) +
  theme_classic() +
  labs(x = "Number of spatial neighbors (k)", y = "Largest-component fraction")
```

A curve that rises more slowly or shifts to the right needs a larger `k` to
reach full connectivity; at any fixed `k`, a lower curve means that a smaller
fraction of the patch belongs to its largest connected component. The opposite
pattern, an earlier rise toward 1, indicates stronger local connectivity.

Interpret this plot alongside assignment coverage:

```{r grid-assignment-summary, class.source = "fold-show"}
assignment_summary_grid[, c(
  "npatches", "n_nonempty_patches", "fraction_assigned",
  "fraction_unassigned"
)]
```

The k-NN graph is built among all epithelial cells supplied to the diagnostic,
before edges are restricted to cells with the same patch label. Consequently,
some of a cell's `k` nearest neighbors may belong to another patch or may be
unassigned; neither type contributes a same-patch edge. If the fraction or
spatial distribution of unassigned cells changes with `npatches`, the
connectivity curves can change partly for that reason. This is why the curves,
assignment coverage, and spatial maps should be considered together.

In this run, the unassigned fractions are all close to 0.4%, whereas the median
curve reaches 1 earliest for `npatches = 150`; the observed median difference
therefore does not indicate worse connectivity at the largest setting.

# Example of filtering

After investigating the diagnostics above, we can use dataset-specific
thresholds and apply them reproducibly. The values below are illustrative choices
for this example.

```{r investigative-filtering, class.source = "fold-show"}
# Dataset-specific thresholds for this example.
min_cells <- 50
min_x_sd <- 0.5
min_stability <- 0.7
min_strict_connectivity <- 0.9

keep <- with(
  patch_data,
  patch[
    n_cells >= min_cells &
      is.finite(x_sd) & x_sd >= min_x_sd &
      !is.na(membership_stability) & membership_stability >= min_stability &
      strict_component_fraction >= min_strict_connectivity
  ]
)
patches_filtered <- patches
patches_filtered[!is.element(patches_filtered, keep)] <- NA
c(patches_retained = length(keep), cells_retained = sum(!is.na(patches_filtered)))
```

# Summary

`getPatchDiagnostics()` summarizes overall assignment coverage and evaluates
each patch's size, within-patch variation, connectivity, and membership
stability. The connectivity curve shows how spatial connectivity changes as
the neighbor graph is widened.

A practical screening workflow is therefore:

1. check `assignment_summary` for unexpectedly low coverage;
2. inspect `n_cells` and `x_sd` together, because a patch needs both enough
   observations and enough predictor variation for a useful within-patch model;
3. use `strict_component_fraction` to flag candidates, then distinguish a local
   split from a persistent one with `min_connectivity_k`, the full
   `connectivity_curve`, and a spatial map;
4. check `membership_stability`; when it is low, consider
   whether more iterations or different construction settings are warranted;
5. compare plausible settings such as `npatches`, then document any thresholds
   and retain the code that applies them.

These diagnostics assess patch construction; they do not by themselves decide
which patches are biologically informative. Final inclusion may also depend on
the expression abundance and variability,
residual diagnostics, tissue anatomy, and the biological question. The
`mouse_colon` vignette demonstrates how retained patches feed into
patch-specific differential expression, residual spatial checks, and
meta-analysis.

# Session info

```{r session-info}
sessionInfo()
```
