---
title: "Space Mosaic"
subtitle: "Benchmarking DE methods"
author:
- name: Patrick Danaher
  affiliation: Bruker Spatial Biology, Seattle
  email: patrick.danaher@bruker.com
- name: Matteo Calgaro
  affiliation: Università degli Studi di Padova
  email: matteo.calgaro@unipd.it
- name: Robert Castelo
  affiliation:
    - &idupf Universitat Pompeu Fabra, Barcelona
  email: robert.castelo@upf.edu
- name: Pere Moles
  affiliation: *idupf
  email: pere.moles@upf.edu
date: "`r BiocStyle::doc_date()`"
package: "`r pkg_ver('SpaceMosaic')`"
vignette: >
  %\VignetteIndexEntry{Space Mosaic}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
  %\VignetteKeywords{GeneExpression, Spatial, Transcriptomics}
output:
  BiocStyle::html_document:
    toc: true
    toc_float: true
    number_sections: true
    fig_captions: yes
---

**License**: `r packageDescription("SpaceMosaic")[["License"]]`

```{r setup, include=FALSE}
options(width=80)
knitr::opts_chunk$set(collapse=TRUE,
                      message=FALSE,
                      warning=FALSE,
                      comment="",
                      fig.align="center",
                      fig.wide=TRUE)
```

# Introduction

Here we illustrate the DE methods available in the SpaceMosaic package. We benchmark these methods using the Cadinu et al. (2024) MERFISH spatial transcriptomics data available from the Bioconductor package `r Biocpkg("MerfishData")`.

```{r load-data, message=FALSE}
library(SpatialExperiment)
library(MerfishData)

spe <- MouseColonIbdCadinu2024()
spe
```

Subset the data to sample 1 slice 1.

```{r}
spe <- spe[, spe$sample_id == "1" & spe$slice_id == "1"]
dim(spe)
```

Calculate patches that split the data into contiguous regions to maximize the
statistical power of detection.

```{r get-patches, message=FALSE}
library(SpaceMosaic)

set.seed(123)  # Set seed for reproducibility
spe <- getPatches(spe, cell_type_column="tier1",
                  response_cell_type="Epithelial",
                  explanatory_cell_type="Immune",
                  mmxpixel=0.000109,
                  npatches=40, n_iters=25)
spe
```

The patches are stored in a reduced dimension slot, named with the convention
`patches_<response_cell_type>_<explanatory_cell_type>`.

```{r}
reducedDimNames(spe)
cellpatches_epi_imm <- reducedDim(spe, "patches_Epithelial_Immune")
cellpatches_epi_imm
```

The metadata of the previous object contains patch-level information.

```{r}
metadata(cellpatches_epi_imm)
```

Visualize the patches.

```{r plot-patches, fig.width=5, fig.height=7, out.width="600px", fig.cap="Patches for Epithelial cells with respect to Immune cells."}
plotPatches(spe, "patches_Epithelial_Immune", with = "ggplot")
```

# DE analysis

## Run DE analysis using the "hasty" method.

```{r de-hasty}
# Get the results and measure the time taken
start <- Sys.time()
de_hasty <- patchDE(spe, "patches_Epithelial_Immune", method = "hasty", verbose = TRUE)
elapsed <- Sys.time() - start

cat("Time taken for hasty DE:", elapsed, "\n")
```

## Run DE analysis using the "limma" method.

```{r de-limma}
# Get the results and measure the time taken
start <- Sys.time()
de_limma <- patchDE(spe, "patches_Epithelial_Immune", method = "limma", verbose = TRUE)
elapsed <- Sys.time() - start 

cat("Time taken for limma DE:", elapsed, "\n")
```
